Parsing Postfix log files

### 2007-07-14 doc/logparser.tex
Wrote prior art section discussing Log Mail Analyser. Documented handling out of order log lines using valid combos. Updated mail coverage section now that valid combos are working and have reduced the number of warnings and mails in the state tables. Some cleanups. Another attempt to break URLs properly.

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
\DeclareUrlCommand\urlwrap{}
\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib,nottoc]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{%
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
    breaklinks  = true,
}

@@

XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS WITH
THE RESULTING MAILS\@; FALSE POSITIVES, FALSE NEGATIVES, ETC\@.

@@

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as database
ready and subsequently entered by the parent mail once it has been
identified.  Quite apart from identifying mail injected in an unknown
fashion, or bugs in the parser, unknown origin mails need to be marked as
such because they lack some data present in the parent mail; they must copy
that data from their parent before being entered in the database.
The effect of this complication upon the algorithm is discussed in
section~\ref{tracking-re-injected-mail}.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\ref{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections.  See
section~\ref{tracking-re-injected-mail} for a full discussion.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~11) ---
        section~\ref{tracking-re-injected-mail} discusses this in detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from memory (flowchart:~15).  For
        re-injected mails see section~\ref{tracking-re-injected-mail}.

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, and that
is explained in section~\ref{tracking-re-injected-mail}.

@@

\label{tracking-re-injected-mail}

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in
section~\ref{tracking-re-injected-mail}, but unlike the solution described
therein bounce notifications do not eventually have a log line which
identifies their source.  Heuristics must be used to identify bounce
notifications, and those heuristics are:

@@

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (all the
complications discussed in section~\ref{additional complications} were
diagnosed and sometimes detected in this way).  The parser issues warnings
when it detects any errors, some of which may alert the user to a problem,
e.g.\ when a queueid is reused before the previous mail is fully dealt
with, when a queueid or pid is not found,\footnote{There will often be a
few warnings about missing queueids or pids in the first few hundred or
thousand log lines because the earlier log lines are in the previous log
file; loading the saved state from the previous log file will solve this
problem} or when there are problems tracking a child mail (see
section~\ref{tracking-re-injected-mail}).  There should be few or no
warnings, and the state table should only contain entries for mails which
had yet to be delivered when the logfiles ended.

At the time of writing the parser is being tested with 93 log files.  There
are 28 warnings produced, but because the parser errs on the side of
producing more warnings rather than fewer those 25 warnings represent only
10 problems:

@@

    \item 3 connections (5 warnings) started before the first log file, so
        their initial log entries are missing.

    \item 5 mails (20 warnings) have out of order log entries, leading to
        warnings about missing data and queueids.  The two log files where
        the out of order log lines occur are more than an order of
        magnitude larger than normal, due to the extremely large volume of
        mail handled by the server on those days; process scheduling
        vagaries and the undefined order of messages sent to syslog(8) may
        explain why the lines were out of order.

The first of these problems cannot be dealt with; although out of order log
lines (the second problem) are dealt with in other circumstances (see
sections~\ref{tracking-re-injected-mail},~\ref{discarding cleanup
lines}~and~\ref{pickup logging after cleanup}), dealing with these lines
would introduce a large amount of complication (deferred warnings, deferred
committal, ambiguity) for a problem which occurs only five times in
60721709 log lines.

The state tables contain entries for mails not yet delivered when the
parser finishes execution.  Ideally all they should contain are mails which
are awaiting delivery after the period covered by the logs, though they may
also contain mails whose initial entries are not contained in the logs.
Any other entries are evidence of a failure in parsing or an aberration in
the logs.  After parsing the 93 test log files the state tables contain 25
entries, breaking down into:

    \item 1 connection which started only seconds before the logs ended
        and had not completed.

@@

    \item 4 mails resulting from the out of order log entries described
        above.

    \item 9 mails whose initial log entries are not present in the logs.

>>>>

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
\DeclareUrlCommand\urlwrap{}
\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib,nottoc]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{%
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}

@@

@@

\subsection{Previous research in this area}

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
quite different from his parser: it attempts to present correlated log data
in a form suitable for a systems administrator to search using grep(1),
producing a CSV file\footnote{Comma-Separated Value, a very simple format
where a record is stored in a single line, with fields separated by a comma
or other punctuation symbol.  Problems with CSV files include the
difficulty in escaping separators, providing multiple values for a field
(e.g.\ multiple recipients), adding new fields, and lack of a bundled schema
describing the fields.} which can subsequently be imported into a MySQL or
Berkeley DB database.  The data stored is limited to time and date,
hostname and IP address of client, sender and recipient addresses, and the
SMTP code.  The decision to support both CSV and Berkeley DB appears to
have been a limiting factor: the values which can be stored are limited,
it's difficult to link different records (foreign keys in SQL enable this
easily), and ad-hoc queries require extensive code to be written.  Berkeley
DB is not an SQL database, it only supports (key, value) pairs; in the main
table the value used is a CSV line, and the key is an integer referred to
by other tables.  The secondary by-sender, by-recipient, by-date and by-IP
tables use the sender/recipient/date/IP as the key, and a CSV list of
integers referring to the main table is used as the value.  In effect this
re-implements SQL foreign keys, but without the full functionality offered
by even the most basic of SQL databases.  It also requires custom code to
search on some combination of the above.

Handling of multiple recipients, SMTP codes or remote servers\footnote{A
single mail may be sent to multiple remote servers if it was addressed to
recipients in different domains, or Postfix needs to try multiple servers
for one or more recipients.} is not explained in the paper, but the options
appear to be:

    \item Discard some of the data; this option is obviously undesirable,
        and presumably was not chosen by the LMA authors.

    \item Combine all data into a CSV line, which is then embedded into the
        main CSV line for that mail, making searching even more difficult
        and requiring a second separating character.

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an SQL database, unlike with Berkeley
DB\@.

Parsing is hand-coded and requires major changes to the code to parse new
log lines or extract additional data.  It does not appear to deal with any
of the complications discussed in
sections~\ref{complications}~and~\ref{additional complications}.  The
program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
a number of warnings and no usable output.

In summary Log Mail Analyser appears to be a proof of concept, written to
demonstrate the point of the paper (that having this information in an
accessible fashion is useful to systems administrators), rather than a
program designed to be extensible and useful in a production environment.

% Literature review notes:
%
% Hard-coded parsing, requiring code changes to add more.  Attempts to
% correlate log lines, saves data to database for data mining purposes.
% Hard to extend/expand/understand.  Appears to only save: date and hour,
% DNS name and IP address host, mail server IP address, sender, receiver
% and e-mail status (sent, rejected).  Undocumented schema.  Design
% decision to use CSV as an intermediate format between the log file and
% the database seems to have been restrictive.  Appears to require a
% queueid but majority of log entries (e.g. rejections) lack a queueid.
% Supports whitelisting IP addresses when parsing logs, but whitelisting
% when generating reports/data mining would be preferable.  Supporting
% Berkeley DB is probably limiting the software - an example is the
% difficulty in searching a pipe-delimited string, so they have
% re-implemented foreign keys with tables keyed by ip address etc. pointing
% at the main table - this also won't scale well.  There doesn't appear to
% be any attempt to deal with the complications I've encountered: their
% parsing isn't detailed enough to encounter them.  It doesn't run
% properly; doesn't create any output; throws up errors.

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as database
ready and subsequently entered by the parent mail once it has been
identified.  Quite apart from identifying mail injected in an unknown
fashion, or bugs in the parser, unknown origin mails need to be marked as
such because they lack some data present in the parent mail; they must copy
that data from their parent before being entered in the database.
The effect of this complication upon the algorithm is discussed in
section~\ref{tracking re-injected mail}.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\ref{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections.  See
section~\ref{tracking re-injected mail} for a full discussion.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~11) ---
        section~\ref{tracking re-injected mail} discusses this in detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from memory (flowchart:~15).  For
        re-injected mails see section~\ref{tracking re-injected mail}.

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, and that
is explained in section~\ref{tracking re-injected mail}.

@@

\label{tracking re-injected mail}

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in
section~\ref{tracking re-injected mail}, but unlike the solution described
therein bounce notifications do not eventually have a log line which
identifies their source.  Heuristics must be used to identify bounce
notifications, and those heuristics are:

@@

\subsubsection{Out of order log lines}

\label{out of order log lines}

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\ref{tracking
re-injected mail},~\ref{discarding cleanup lines}~and~\ref{pickup logging
after cleanup}.  In the 93 log files used for testing this occurs only five
times in 60721709 log lines, but for completeness of the algorithm it
should be dealt with.  The five occurrences in the test log files have the
same characteristics: the postfix/local log line showing delivery to a
local mailbox occurs after the postfix/qmgr log line showing removal of the
mail from the queue because delivery is completed.  This causes problems:
the mail is not complete, so entry into the database fails; a new mail is
created when the postfix/local line is parsed and remains in the state
tables; four warnings are issued per pair of out of order log lines.

The solution to this problem is to examine the list of programs which have
logged messages for each mail, comparing the list against a table of
known-good combinations of programs.  If combination is found in the valid
list the mail can be entered in the database; if the combination is not
found entry must be postponed and the mail flagged.  The SAVE\_BY\_QUEUEID
action checks for the flag and retries entry; if the additional log lines
have caused the mail to reach a valid combination entry will succeed,
otherwise it must be postponed once more.

The list of valid combinations is given below.  Every mail will
additionally have log entries from postfix/cleanup and postfix/qmgr; any
mail may also have log entries from postfix/bounce, postfix/postsuper, or
both.

    \item [postfix/local:] Local delivery of bounce notification, or local
        delivery of forwarded/tracked mail.

    \item [postfix/local, postfix/pickup:] Locally submitted mail, locally
        delivered.

    \item [postfix/local, postfix/pickup, postfix/smtp:] Locally submitted
        mail, \newline both local and remote delivery.

    \item [postfix/local, postfix/smtp, postfix/smtpd:] Mail accepted from
        a remote client, both local and remote delivery.

    \item [postfix/local, postfix/smtpd:] Mail accepted from a remote
        client, local delivery.

    \item [postfix/pickup, postfix/smtp:] Locally submitted mail, remote
        delivery.

    \item [postfix/smtp:] Remote delivery of forwarded/tracked mail.

    \item [postfix/smtp, postfix/smtpd:] Mail accepted from a remote
        client, remotely delivered (relaying mail for clients on the local
        network).

This applies to accepted mails only, not to rejected mails.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (all the
complications discussed in section~\ref{additional complications} were
diagnosed and sometimes detected in this way).  The parser issues warnings
when it detects any errors, some of which may alert the user to a problem,
e.g.\ when a queueid is reused before the previous mail is fully dealt
with, when a queueid or pid is not found,\footnote{There will often be a
few warnings about missing queueids or pids in the first few hundred or
thousand log lines because the earlier log lines are in the previous log
file; loading the saved state from the previous log file will solve this
problem} or when there are problems tracking a child mail (see
section~\ref{tracking re-injected mail}).  There should be few or no
warnings, and the state table should only contain entries for mails which
had yet to be delivered when the logfiles ended.

At the time of writing the parser is being tested with 93 log files.  There
are 5 warnings produced, but because the parser errs on the side of
producing more warnings rather than fewer those 5 warnings represent only 3
instances of 1 problem: 3 connections started before the first log file, so
their initial log entries are missing, leading to warnings when their log
lines are parsed.

The state tables contain entries for mails not yet delivered when the
parser finishes execution.  Ideally all they should contain are mails which
are awaiting delivery after the period covered by the logs, though they may
also contain mails whose initial entries are not contained in the logs.
Any other entries are evidence of a failure in parsing or an aberration in
the logs.  After parsing the 93 test log files the state tables contain 22
entries, breaking down into:

@@

    \item 2 connections which started only seconds before the logs ended
        and had not completed.

    \item 9 mails whose initial log entries were not present in the logs.

@@
